
---
title: "Divorces in New Zealand"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Divorces in New Zealand}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

# Introduction

In this vignette we use **BayesRates** to calculate age-specific rates and turn these age-specific rates into total divorce rates. We use administrative data on divorce in New Zealand in New Zealand. The counts of divorce and population in the dataset are large, so sampling errors are small.


# Preliminaries

## Packages

Besides **BayesRates** we load **dplyr**, for data manipulation, and **ggplot** for plotting.

```{r setup}
library(BayesRates)
library(dplyr)
library(ggplot2)
```

## Data 

The data on divorces is ins `nz_divorces`, a data frame contained in the **BayesRates** package. `nz_divorces has four columns:

- `age`: a factor with levels `"15-19"`,`"20-24"`,...,`"60-64"`,`"65+"`;  
- `sex`: a character vector with values `"Female"` and `"Male"`;  
- `time`: an integer vector with values `1992`,...,`2021`;  and
- `nevent`: a numeric vector with the number of divorces.

```{r}
nz_divorces
```

The data on population is in `nz_population`, also contained in the **BayesRates** package. `nz_population` has with four columns:

- `age`: a factor with levels `"15-19"`,`"20-24"`,...,`"60-64"`,`"65+"`;  
- `sex`: a character vector with values `"Female"`, `"Male"`;  
- `time`: an integer vector with values being `1992`,...,`2021`;  and
- `py`: a numeric vector with person-years of exposure.

```{r}
nz_population
```

The person-years of exposure, and hence the number of divorces, are both quite large.
```{r}
summary(nz_population$py)
summary(nz_divorces$nevent)
```

# Rates in a single period, both sexes combined

We start with the relatively simple where we are estimating age-specific rates in a single period, not distinguishing between females and males.

As a first step, we create two new datasets each only containing data for 2020.

```{r}
nz_divorce_2020 <- nz_divorces |>
  filter(time == 2020) |>
  select(-time)

nz_popn_2020 <- nz_population |>
  filter(time == 2020) |>
  select(-time)
```

We use the `smooth_age()` function to smooth the observed across age. 
```{r}
res_age <- smooth_age(nevent_df = nz_divorce_2020,
                      py_df = nz_popn_2020,
                      age_width_df = nz_age_width_df,
                      age_min = 15)
```

In the call to `smooth_age()`, the parameter `nevent_df` specifies the data frame for numbers of events, and the parameter `py_df` specifies the data frame for person years of exposure. The parameters `age_width_df` and `age_min` provide information about age groups. `nz_age_width_df` is an existing data frame in the **BayesRates** package that gives the width of each age group.

```{r}
head(nz_age_width_df)
```

We have not told `smooth_age()` to calculate separate rates for females and males, so, before fitting, it aggregates events and person-years across the sexes.

A summary of the model specification can be obtained by printing the return value from `smooth_age()`.

```{r}
res_age
```

The number of events is modeled by a Poisson distribution with mean equal to an underlying rate multiplied by person-years of exposure.  The logarithm of the underlying rate equals the age effect, which is modeled by a spline model.  The model for the age effect is `Spline()`.  The value for `agevar` is `"age"`, indicating that variable with information on age groups is called `"age"`.  The value of `byvar` is empty, indicating that we fit a single model, within everyone belonging to the same group.  The value of `ndraw` is 1000, indicating that the number of posterior draws is 1000, the default.

We use the `components()` function to extract model-based estimates of rates disaggregated by age.

```{r}
components(res_age, what = "rates")
```

`components()` returns a data frame. In this data frame,

- `age.mid` gives the middle point for each age group (which is helpful for plotting);
- `.fitted` gives a point estimate of the rate;
- `.lower` and `.upper` give lower and upper bounds of the credible interval of the rate (a 95% credible interval by default); and
- `.probability` gives a list containing the posterior draws of the rate.


`components()` can also be used to extract model-based estimates of age effects:

```{r}
components(res_age, what = "age_effect")
```

Function `augment()` returns the original data plus model-based estimates of rates.
```{r}
rates_age_df <- augment(res_age)
rates_age_df
```
Since the model does not distinguish sexes, the values of '.fitted', '.lower', '.upper' and '.probability' are the same for females and males at the each age.

We plot the model-based estimates of rates (including point estimates and credible intervals) and the observed rates (in blue), separately for each sex.
```{r}
ggplot(rates_age_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```

Function `total_rate()` sums across ages to give total rates. 
```{r}
total_rate(res_age)
```

# Rates in a single period, separating sexes

We next fit a separate model for each sex.  To do this, we set the argument `byvar` in `smooth_age()` to `"sex"`.
```{r}
res_age_bysex <- smooth_age(nevent_df = nz_divorce_2020,
                            py_df = nz_popn_2020,
                            age_width_df = nz_age_width_df,
                            age_min = 15,
                            byvar = "sex")
res_age_bysex
```

The rates returned by `augment()` now distinguish between sexes

```{r}
rates_age_bysex_df <- augment(res_age_bysex)
rates_age_bysex_df
```

Graphing the results shows that the model now recognizes the female-male differences in age patterns.
```{r}
ggplot(rates_age_bysex_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```

Function `total_rate()` now produces separate results for females and males.

```{r}
total_rate(res_age_bysex)
```

# Alternative specifications of age effect

The `spec_age` argument to `smooth_age()` specifies the prior model to be applied to age effects. The default value for `spec_age` is `Spline()`. In this section we experiment with alternative values.

## Modifying the spline model for age effects

When `spec_age` is set to `Spline()`, age effects are modelled using a penalised spline, with degrees of freedom equal to `max(ceiling(0.7 * n), 4)`, where `n` is the number of age groups.  In our data, `n` equals 11, so the degrees of freedom is 8.

Reducing the degrees of freedom used within `Spline()` may lead smoother age effects. Here we set the degrees of freedom to `5`, and graph the results.

```{r}
res_age_bysex_5df <- smooth_age(nevent_df = nz_divorce_2020,
                                py_df = nz_popn_2020,
                                age_width_df = nz_age_width_df,
                                age_min = 15,
                                byvar = "sex",
                                spec_age = Spline(df = 5))

rates_age_bysex_5df <- augment(res_age_bysex_5df)

ggplot(rates_age_bysex_5df,
        aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```

The lower degrees of freedom does indeed increase the smoothing, and the modelled estimates are further away from the direct estimates.


## Second-order random walk model for age

We can also use a second order random walk model to smooth the observed rates across age for each sex.

```{r}
res_age_bysex_rw <- smooth_age(nevent_df = nz_divorce_2020,
                               py_df = nz_popn_2020,
                               age_width_df = nz_age_width_df,
                               age_min = 15,
                               byvar = "sex",
                               spec_age = RW2())

rates_age_bysex_rw <- augment(res_age_bysex_rw)

ggplot(rates_age_bysex_rw,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```


# Smoothing over age and time

We now return to the full dataset, and apply function `smooth_agetime()`, which simultaneously smooths over age and time.

## Fixed age pattern

In our first model, we asssume that age-patterns are constant over time. We set the parameter `spec_time` to `TimeFixed()`.

```{r}
res_agetime_fix <- smooth_agetime(nevent_df = nz_divorces,
                                  py_df = nz_population,
                                  age_width_df = nz_age_width_df,
                                  age_min = 15,
                                  byvar = "sex",
   		                  spec_time = TimeFixed())
res_agetime_fix
```
The model outputs now have a time dimension.

```{r}
rates_agetime_fix <- augment(res_agetime_fix)
rates_agetime_fix
```

When we graph the results for the period 1992--2021, we see systematic differences between the modelled and direct estimates.

```{r, fig.height = 7}
ggplot(rates_agetime_fix, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = sex),
              alpha = 0.6) +
  geom_line(aes(y = .fitted, color = sex)) +
  geom_point(aes(y = .observed, color = sex),
             size = 0.5)
```

These results suggest that we may need to allow the age-pattern to vary over time.

## Time-varying age pattern

To allow the age-pattern to vary over time, we set `spec_time` to `TimeVarying()`, which is in fact the default. In the `TimeVarying()` model, each age group follows a different random walk, with the random walks for neighbouring age groups being more highly correlated than random walks for distant age groups.

```{r}
res_agetime_vary <- smooth_agetime(nevent_df = nz_divorces,
                                   py_df = nz_population,
                                   age_width_df = nz_age_width_df,
                                   age_min = 15,
                                   byvar = "sex")
res_agetime_vary
```

In this revised model, the systematic differences between modelled and direct estimates disappear.

```{r, fig.height = 7}
rates_agetime_vary <- augment(res_agetime_vary)

ggplot(rates_agetime_vary, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = sex),
              alpha = 0.6) +
  geom_line(aes(y = .fitted, color = sex)) +
  geom_point(aes(y = .observed, color = sex),
             size = 0.5)
```

Using the `components()` function, we can extract the age and age-time effects.

The age effect is an average age pattern over the whole period.

```{r}
age_effect <- components(res_agetime_vary, what = "age_effect")

ggplot(age_effect,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line()
```

The age-time effect shows departures from the average effect in each period

```{r, fig.height = 7}
time_effect <- components(res_agetime_vary, what = "time_effect") 

ggplot(time_effect, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = sex),
              alpha = 0.6) +
  geom_line(aes(y = .fitted, color = sex))
```

The age-time effects rotate in an anti-clockwise direction. This implies that the overall shellage-pattern is become more skewed towards older ages over time.


Finally, we calculate time-varying total rates.
```{r}
total_varying <- total_rate(res_agetime_vary)

ggplot(total_varying, 
       aes(x = time, y = .fitted, ymin = .lower, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```
