---
title: "vig1_nz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vig1_nz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

In this example, we have the number of divorce events and person-years of exposure for New Zealand by age groups (15-19, 20-24, ..., 65+), sex, and year (1992, ..., 2021).  The values of the person-years of exposure are quite large and thus sampling error is not very large. 

```{r setup}
library(BayesRates)
library(dplyr)
library(ggplot2)
```

'nz_divorces' is an existing data frame in the *BayesRates* package with four columns:  
(1) 'age': a factor variable with levels being "15-19","20-24",...,"60-64","65+";  
(2) 'sex': a character variable with possible values being "Female" and "Male";  
(3) 'time': an integer variable with possible values being 1992,...,2021;  
(4) 'nevent': a numeric variable recording the number of divorce events.

```{r}
head(nz_divorces)
```

'nz_population' is an existing data frame in the *BayesRates* package with four columns:  
(1) 'age': a factor variable with levels being "15-19","20-24",...,"60-64","65+";  
(2) 'sex': a character variable with possible values being "Female", "Male";  
(3) 'time': an integer variable with possible values being 1992,...,2021;  
(4) 'py': a numeric variable recording person-years of exposure.
```{r}
head(nz_population)
```

The person-years of exposure are quite large, and correspondingly the numbers of divorces are not small.
```{r}
summary(nz_population["py"])
table(nz_divorces["nevent"])
```

We keep the subsets of 'nz_divorces' and 'nz_population' for 2020, and remove the 'time' variable.  To do this, we use the '%>%' operator from the 'dplyr' package that turns 'x %>% f(y)' into 'f(x,y)' so that the result from one step is piped into the next step.
```{r}
nz_divorce_2020 <- nz_divorces %>%
  filter(time == 2020) %>%
  select(-time)

nz_popn_2020 <- nz_population %>%
  filter(time == 2020) %>%
  select(-time)
```

We use the *smooth_age* function to smooth the observed unisex rates across age. The parameter *nevent_df* specifies the data frame for numbers of events, and the parameter *py_df* specifies the data frame for person years of exposure.  When we work with age groups, we also need to specify the parameter *age_width_df* which is a data frame recording the width of each age group, and the parameter *age_min* which is the starting age.  As input to the parameter *age_width_df*, we use 'nz_age_width_df', an existing data frame in the *BayesRates* package that gives the width of each age group.
```{r}
head(nz_age_width_df)

res_age <- smooth_age(nevent_df = nz_divorce_2020,
                      py_df = nz_popn_2020,
                      age_width_df = nz_age_width_df,
                      age_min = 15)
```
Inside the *smooth_age* function, numbers of events and person-years of exposure for females and males are aggregated. The default model for the age effects on log rates is a spline model with degrees of freedom equal to max(ceiling(0.7 * n), 4), where n is the number of age groups.  Here we have n=11, hence the degrees of freedom is 8.

We print the model specification as follows.
```{r}
res_age
```
The number of events is modeled by a Poisson distribution with mean equal to an underlying rate multiplied by person-years of exposure.  The logarithm of the underlying rate equals the age effect, which is modeled by a spline model.  The model for the age effect is *Spline()*.  The value of 'agevar' is 'age', indicating that the name of the variable that records age is 'age'.  The value of 'byvar' is empty, indicating that we fit a single model (not classified by any variable).  The value of 'ndraw' is 1000, indicating that the number of posterior draws is 1000 (default value).

We use the *components* function to extract model-based estimates of rates disaggregated by age.
```{r}
components(res_age, what = "rates")
```
In the returned data frame, the variable 'age.mid' gives the middle point for each age group, the variable '.fitted' gives the point estimate of the rate, the variables '.lower' and '.upper' give the lower and upper bounds of the credible interval of the rate (95% credible interval by default), and the variable '.probability' gives a list containing the posterior draws of the rate.  

We also use the *components* function to extract model-based estimates of age effects.
```{r}
components(res_age, what = "age_effect")
```

We use the *augment* function to combine the observed data with the model-based estimates of rates.
```{r}
rates_age_df <- augment(res_age)
head(rates_age_df)
```
Since the model has been constructed for unisex rate, the values of '.fitted', '.lower', '.upper' and '.probability' are the same for males and females at the same age.

We plot the model-based estimates of rates (including point estimates and credible intervals) and the observed rates (in blue), separately for each sex.
```{r}
ggplot(rates_age_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```

We use the *total_rate* function to obtain estimates of the unisex total rates. 
```{r}
total_rate(res_age)
```

We next fit a separate model for each sex.  To do this, we set the parameter *byvar* in the *smooth_age* function to "sex".
```{r}
res_age_bysex <- smooth_age(nevent_df = nz_divorce_2020,
                      py_df = nz_popn_2020,
                      age_width_df = nz_age_width_df,
                      age_min = 15,
                      byvar = "sex")
res_age_bysex
```

We combine observed data with model-based estimates of rates, plot the results, and obtain estimates of total rates.
```{r}
rates_age_bysex_df <- augment(res_age_bysex)
head(rates_age_bysex_df)

ggplot(rates_age_bysex_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")

total_rate(res_age_bysex)
```

We can also change the model specification.  For example, we can use a spline model with 9 degrees of freedom to smooth the observed rates across age for each sex.
```{r}
res_age_bysex2 <- smooth_age(nevent_df = nz_divorce_2020,
                      py_df = nz_popn_2020,
                      age_width_df = nz_age_width_df,
                      age_min = 15,
                      byvar = "sex",
                      spec_age = Spline(df = 9))
res_age_bysex2

rates_age_bysex2_df <- augment(res_age_bysex2)
head(rates_age_bysex2_df)

ggplot(rates_age_bysex2_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")

total_rate(res_age_bysex2)
```

We can also use a second order random walk model to smooth the observed rates across age for each sex.
```{r}
res_age_bysex3 <- smooth_age(nevent_df = nz_divorce_2020,
                      py_df = nz_popn_2020,
                      age_width_df = nz_age_width_df,
                      age_min = 15,
                      byvar = "sex",
                      spec_age = RW2())
res_age_bysex3

rates_age_bysex3_df <- augment(res_age_bysex3)
head(rates_age_bysex3_df)

ggplot(rates_age_bysex3_df,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")

total_rate(res_age_bysex3)
```

Next, for each sex, we use data for all years and smooth the observed rates across age and time.  By default, we model the age effects on log rates using a spline model with degrees of freedom equal to max(ceiling(0.7 * n), 4), where n is the number of age groups, and allow the time effects on log rates to vary for different age groups.  A correlated first-order random walk is used to model the time effects, where the correlation is higher for age groups that are closer. 
```{r}
res_agetime_bysex <- smooth_agetime(nevent_df = nz_divorces,
                              py_df = nz_population,
                              age_width_df = nz_age_width_df,
                              age_min = 15,
                              byvar = "sex")
res_agetime_bysex
```
In the model specification, the logarithm of the underlying rate equals the age effect plus the time effect.  The model for the time effect is *TimeVarying()*.  The value of 'timevar' is 'time', indicating that the name of the variable that records time is 'time'.  The value of 'byvar' is 'sex', indicating that a model is fitted separately for each sex.

We use the *augment* function to combine data and model-based estimates of rates, and plot the results.
```{r}
rates_agetime_bysex_df <- augment(res_agetime_bysex)
head(rates_agetime_bysex_df)
```

```{r, fig.height = 7}
ggplot(rates_agetime_bysex_df, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = sex),
              alpha = 0.6) +
  geom_line(aes(y = .fitted, color = sex)) +
  geom_point(aes(y = .observed, color = sex),
             size = 0.5)
```

We use the *components* function to extract the model-based estimates of age effects, and plot them.
```{r}
age_effect <- components(res_agetime_bysex, what = "age_effect")
head(age_effect)

ggplot(age_effect,
       aes(x = age.mid, ymin = .lower, y = .fitted, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line()
```

We use the *components* function to extract the model-based estimates of time effects, and plot them.
```{r}
time_effect <- components(res_agetime_bysex, what = "time_effect") 
head(time_effect)
```

```{r, fig.height = 7}
ggplot(time_effect, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = sex),
              alpha = 0.6) +
  geom_line(aes(y = .fitted, color = sex))
```

We obtain estimates of the total rates and plot them.
```{r}
total_rate = total_rate(res_agetime_bysex)

total_rate

ggplot(total_rate, 
       aes(x = time, y = .fitted, ymin = .lower, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed), col = "blue")
```
