---
title: "Divorces in China"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Divorces in China}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

# Introduction

In this vignette, we use **BayesRates** to calculate smoothed age-specific divorce rates in China. The input data come from household surveys, and sample sizes are small.

## Preliminaries

## Packages

We load **BayesRates** and the two tidyverse packages **dplyr** and **ggplot**.

```{r setup}
library(BayesRates)
library(dplyr)
library(ggplot2)
```


## Data

Our data on divorces is `cn_divorces`, a data frame that is part of the **BayesRates** package. `cn_divorces has four columns:

- `age`: an integer vector with values 15,...,64;  
- `sex`: a character vector with values "Female" and "Male";  
- `time`: an integer vector with values 1980,...,2018;  and 
- `nevent`: a numeric vector with the number of divorces.

```{r}
cn_divorces
```

`cn_population` contains data on population, and is also part of the **BayesRates**. Its columsn are:

- `age`: an integer vector with values 15,...,64;  
- `sex`: a character vector with values "Female" and "Male";  
- `time`: an integer vector with values 1980,...,2018;  and 
- `py`: a numeric vector with person-years of exposure.

```{r}
cn_population
```

The person-years of exposure are quite small, and over 70 percent of divorce counts equal zero.

```{r}
summary(cn_population$py)
summary(cn_divorces$nevent)
mean(cn_divorces$nevent == 0)
```

# Fitting a model

We use  function `smooth_agetime()` to smooth the observed rates across age and time, separately for each sex.
```{r}
res <- smooth_agetime(nevent_df = cn_divorces,
                      py_df = cn_population,
                      spec_time = TimeFixed(),
                      byvar = "sex")
```

- `nevent_df` is the data frame for numbers of events
- `py_df` is the data frame for person years of exposure
- `spec_time` is the model used for age-time interactions. Given the limited data, we use a simple model in which the overall level changes with time but the age-pattern remains constant. 

By default, `smooth_agetime()` models age effects (as opposed to age-time interactions) using function `Spline()`, which uses penalised splines to smooth over age groups. The degree of smoothness in `Spline()` can be controlled via function `df`.

A summary of the model can be obtained by printing the model object.

```{r}
res
```

- The number of events is modeled by a Poisson distribution with mean equal to an underlying rate multiplied by person-years of exposure
- The logarithm of the underlying rate equals the age effect plus the time effect
- The model for the age effect is *Spline()*
- The model for the time effect is *TimeFixed()*
- `agevar` is `"age"`. **BayesRates** is assuming that the variable called `"age"` identifies age groups
- `timevar` is `"time"`. **BayesRates** is assuming that the variable called `"time"` identifies time periods
- `byvar` is `"sex"` The model is fitted separately for each sex
- `ndraw` is 1000. By default functions such as `augment()` and `components()` (see below) produce 1000 draws from the posterior distribution.


# Extracting results

Function `augment()` returns a data frame the combines the input data model-based estimates of rates.

```{r}
rates <- augment(res)
rates
```

In the return value from `augment()`

- `age.mid` is the middle point for each single year of age;
- `.fitted` is the point estimate of the rat
- `.lower`, `.upper` are the lower and upper bounds of 95% credible intervals for the rates; and
- `.probability` is a list containing the posterior draws of the rate.  

We extract and graph the rates for females.

```{r, fig.height = 7}
rates_female <- rates |>
  filter(sex == "Female")
  
ggplot(rates_female, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper),
              fill = "salmon") +
  geom_line(aes(y = .fitted),
            color = "darkred") +
  geom_point(aes(y = .observed),
             color = "blue",
	     size = 0.2)
```

# Total rates

Function `total_rate()` sums across age groups to give the total divorce rate.

```{r}
total_rate <- total_rate(res)
total_rate
```

We plot model-based (orange) and direct (blue) estimates of the total divorce rate in China.

```{r}
ggplot(total_rate, 
       aes(x = time, y = .fitted, ymin = .lower, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line() +
  geom_point(aes(y = .observed),
             col = "blue")
```


# Imputing missing values

It is common, in demographic data, to have missing values for some years. **BayesRates** automatically imputes values for years with missing data.

To illustrate, we first create datasets where data is only available for every fifth year.

```{r}
cn_divorces_5 <- cn_divorces |>
  filter(time %in% seq(1980, 2018, 5))
cn_population_5 <- cn_population |>
  filter(time %in% seq(1980, 2018, 5))
```

We use `smooth_agetime()` to smooth rates across age and time, separately for each sex.
```{r}
res_5 <- smooth_agetime(nevent_df = cn_divorces_5,
                        py_df = cn_population_5,
                        spec_time = TimeFixed(),
                        byvar = "sex")
res_5
```

When we use `augment()` function, we only obtain estimates for the observed years.

```{r}
rates_5 <- augment(res_5)
rates_5
```

However, when we use `components()` to extract model-based estimates of rates, we also obtain estimates for the gap years.
```{r}
rates_5_com <- components(res_5, what = "rates")
rates_5_com
```

```{r, fig.height = 7}
rates_5_com_female <- rates_5_com |>
  filter(sex == "Female") 
  
ggplot(rates_5_com_female, aes(x = age.mid)) +
  facet_wrap(vars(time)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper),
              fill = "salmon") +
  geom_line(aes(y = .fitted),
            color = "darkred")
```

`total_rate()` also produces estimates for all years, including ones without data.

```{r}
total_rate_5 <- total_rate(res_5)
```

Credible intervals are wider in years without data.

```{r}
ggplot(total_rate_5, 
       aes(x = time, y = .fitted, ymin = .lower, ymax = .upper)) +
  facet_wrap(vars(sex)) +
  geom_ribbon(fill = "salmon") +
  geom_line(col = "darkred") +
  geom_point(aes(y = .observed),
             col = "blue")
```
